Deciphering enhancer rules underlying early human neuronal and neural crest development.

### 2025-04-08 manuscript.tex
C.M. comments

\author[1,2,3,\orcidlink{0000-0001-7907-1247}]{Seppe De Winter}
\author[1,2,3,\orcidlink{0000-0002-1318-2603}]{Camiel Mannens}
\author[1,2,3]{Valerie Christiaens}
\author[1,2,3,*,\orcidlink{0000-0002-8006-0315}]{Stein Aerts}

@@

\affil[1]{VIB Center of AI \& computational Biology, Leuven, Belgium}
\affil[2]{VIB-KU Leuven Center for Brain and Disease Research, Leuven, Belgium}
\affil[3]{KU Leuven Department of Human Genetics, Leuven, Belgium}

@@

Organoids are a reductionistic biological model system that can be used to interrogate how cell types are encoded in the
genome. Especially organoids derived from pluripotent stem cells are interesting in this context, given that they
demonstrate how a single genome can generate multiple distinct cell types\cite{mccauleyPluripotentStemCellderived2017,
takebeOrganoidsDesign2019, hoferEngineeringOrganoids2021}. Organoids can furthermore, give insights into how genomes
interpret both cell intrinsic and extrinsic signals. For example, epigentic
modifications\cite{zenkSinglecellEpigenomicReconstruction2024}, transcription factor (TF)
perturbations\cite{fleckInferringPerturbingCell2023} or morphogen
signaling\cite{sanchis-callejaDecodingMorphogenPatterning2024}. Another benefit of organoids is that they can be used to
model biological systems that are otherwise - due to ethical and/or practical reasons - inaccessible for study. For
example, early human development.\par

@@

Current efforts are ongoing to determine the breadth of human cell types that can be generated using organoid protocols and how they compare to their primary counterparts, at a transcriptomic level. For example, for brain organoids\cite{heIntegratedTranscriptomicCell2024} and endoderm-derived organoids\cite{xuIntegratedTranscriptomicCell2023}. However, such comparisons at the level of the enhancer-code are still limited\cite{zenkSinglecellEpigenomicReconstruction2024, fleckInferringPerturbingCell2023}. In this study, we will address this gap by determining the enhancer-codes of cell types/states generated by human neural tube organoids, a model system for early human neuronal and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021} and compare these codes to those of cell types/states of the head of a four weeks post conception human embryo.\par

During vertebrate development both the neural tube and neural crest originate from the neural ectoderm. The neural tube
consist of proliferating neuronal progenitors that, through opposing gradients of SHH and BMP/Wnt signaling, have different identities along the dorso-ventral (DV) axis of the embryo later giving rise to distinct neuronal subtypes of the central nervous system\cite{sagnerEstablishingNeuronalDiversity2019, delasRepressiveInteractionsGene2020}. Neural crest cells originate from the neural plate border, through intermediate signals of BMP and Wnt\cite{simoes-costaEstablishingNeuralCrest2015}, and migrate throughout the body giving rise to a multitude of cell types including the peripheral nervous system, craniofacial skeleton and melanocytes\cite{simoes-costaEstablishingNeuralCrest2015}. The gene regulatory networks underlying both DV patterning of neuronal progenitors\cite{sagnerEstablishingNeuronalDiversity2019, delasRepressiveInteractionsGene2020} and neural crest induction, specification, migration and differentiation\cite{simoes-costaEstablishingNeuralCrest2015, candido-ferreiraMultilayeredTranscriptionalControl2023} are already well studied. Furthermore, studies using chromatin accessibility profiling methods at the bulk and single cell level, profiling enhancers encoding cell types and states of neuronal progenitors and neural crest cells are starting to emerge\cite{delasDevelopmentalCellFate2023, zhangCisregulatoryLogicIntegrating2024, huEndothelialRegulatoryCircuits2025} but a detailed descriptions of the enhancer-code of each individual cell types are still limited.\par

@@

To evaluate the biological relevance of the cell types and states obtained from the organoid, we performed 10x single

@@

clear that both early differentiating peripheral neurons and to a greater extend early differentiating neurons of the

@@

@@

@@

$DeepNeuralTube_{embryo}$ was able to distinguish the different AP domains but grouped topics of similar domains
together (i.e., Topic 59, 31 and 62; Topic 52 and 71), resulting in four distinct AP domains (\textbf{Fig.4 c-d}). By
explaining the predictions of $DeepNeuralTube_{embryo}$ on the regions of those domains we found four distinct
homeodomain TF binding motifs, one for each domain (\textbf{Fig. 4e}). From anterior to posterior this is \textit{TAATTA}, \textit{GGATTA}, \textit{TCATGWTGANTGA} and \textit{CATYMATCA}. Notably, the patterns corresponding to the two posterior most topics correspond to dimer-motifs. Previously it was already hypothesized that genome-wide specific binding of homeodomain TFs is accomplished through cooperative binding of two TFs\cite{mannChapter3Hox2009}. Indeed, the presence of instances of the \textit{TCATGWTGANTGA} motif correlates with the expression of both \textit{PAX5/8} and \textit{EN1/2} and the prescence of instances of the \textit{CATYMATCA} motif correlates with the expression of \textit{MEIS1/2}, \textit{PBX1/2} and \textit{HOXB3} (\textbf{Fig. 4d-e} and Supplementary Fig. S4). In contrast, the patterns corresponding to the two anterior most topics correspond to monomer motifs. The presence of instances of these patterns correlates with the expression of \textit{LHX2/9}, \textit{RAX} and \textit{SIX3/6} for the most anterior topic and with the expression of \textit{OTX1/2}, \textit{EMX2}, \textit{DMBX1} and \textit{LMX1A} for the second most anterior topic (\textbf{Fig. 4d-e} and Supplementary Fig. S4).

Focusing on the two most anterior domains, two similar homeodomain TF binding motifs seem to underlie the differential
accessibility of those domains. Both have the core homeobox motif: \textit{TAAT} but one is flanked by the nucleotides \textit{TAG} while the other is flanked by the nucleotides \textit{CCCT}. We were surprised that this slight difference in the two motifs would be enough for encoding two separate AP domains. To test this hypothesis, we first did a substitution experiment where we replaced all instances of \textit{CTAAT\underline{TAG}} with \textit{CTAAT\underline{CCCT}} in regions of the most anterior domain and vice versa (\textbf{Fig. 4f}). Indeed, $DeepNeuralTube_{embryo}$ predicts an anterior-to-posterior shift of accessibility of those regions after the \textit{CTAAT\underline{TAG}} to \textit{CTAAT\underline{CCCT}} substitution and vice versa a posterior-to-anterior shift of accessibility after the \textit{CTAAT\underline{TAG}} with \textit{CTAAT\underline{CCCT}} to  \textit{CTAAT\underline{TAG}} substitution (\textbf{Fig. 4f}). Next, we designed enhancers from scratch by embedding either the  \textit{CTAAT\underline{TAG}} or \textit{CTAAT\underline{CCCT}} motif\cite{taskiranCelltypedirectedDesignSynthetic2024, kempynckCREstedModelingGenomic2025} and using respectively Topic 61 and Topic 31 as target class (\textbf{Fig. 4g}). Embedding of two instances of either motif was sufficient to reach predicted chromatin accessibility levels of equivalent ranges to the predicted chromatin accessibility levels of genomic regions (\textbf{Fig. 4g}). Next, to eliminate potential model bias we repeated the same experiment but optimized for the opposite class. This is, we optimized for Topic 31 by embedding the \textit{CTAAT\underline{TAG}} motif and we optimized for Topic 61 by embedding the \textit{CTAAT\underline{CCCT}} motif (\textbf{Fig. 4h}). In this case, the synthetic design failed to converge for the target class but instead regions with high prediction score for the AP domain that corresponds to the motif used for embedding were generated. This is embedding of the \textit{CTAAT\underline{TAG}} motif still generated regions with high prediction score for Topic 61 and embedding of the \textit{CTAAT\underline{CCCT}} motif generated regions with high prediction score for Topic 31 (\textbf{Fig. 4h}).\par

@@

Focusing on neural crest cells, we identified topics representing cells and regions of both pre- and migratory neural
crest as well as facial mesenchyme (\textbf{Fig. 5a}). Similar to the neuronal progenitors, we used
$DeepNeuralTube_{oranoid}$ and $DeepNeuralTube_{embryo}$ to explain the top 1,000 regions for each organoid and embryo
topic and generated a code-table based on the identified pattern instances (\textbf{Fig. 5b}). Again, both
$DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ found similar patterns and the importance of the instances fo
these patterns is highly correlated across both models (0.73 average Pearson correlation coefficient; \textbf{Fig. 5b}).\par

@@

The migratory neural crest cell state is represented by two organoid topics (Topic 65 and 59) and a single embryo topic (Topic 94). Nuclear receptor, AP2, ZIC (only in organoid topics) and SOX-dimer (organoid Topic 59 but not 65) are important for this state (\textbf{Fig. 5b-c}). We could not link the nuclear receptor instances to a specific TF based on the correlation of expression and the importance score of these instances. The importance of AP2 instances correlates with the expression of \textit{TFAP2A}, \textit{TFAP2B} and \textit{TFAP2C} (\textbf{Fig. 5d}), ZIC with \textit{ZIC1} to \textit{ZIC5} and the SOX-dimer with expression of \textit{SOX10} (\textbf{Fig. 5d} and Supplementary Fig. S5). The switch from TFAP2A/C in pre-migratory neural crest to TFAP2A/B in migratory neural crest is consistent with a previously reported function of both heterodimers in respectively neural plate border induction and neural crest specification\cite{rothsteinHeterodimerizationTFAP2Pioneer2020}. After the pre-migratory and before the migratory neural crest state another topic is found in both organoid (Topic 60) and embryo (Topic 105) for which instances of RFX are important (\textbf{Fig. 5b-c}) that correlate with the expression of \textit{RFX3/4} (\textbf{Fig. 5d} and Supplementary Fig. S5).\par

@@

Next, we investigated which patterns co-occur in the same genomic regions. For this purpose, we calculated the Jaccard
index of genomic regions based on the presence of an instance of each identified pattern (\textbf{Fig. 5e}). Using this
method we found three sets of regions that are predicted to contain instances of different set of TFs. One set where
instances of TEAD, GRHL and ZEB co-occur, one where instances of Nuclear receptor, AP2 and ZIC
instances co-occur and one set where instances of FOX and TWIST1/homeodomain co-occur.\par

@@

For both the organoid and embryo we found two topics (organoid Topic 6 and 4 and embryo Topic 10 and 18) for which
instances of patterns from neuronal progenitors, more precisely the floorplate, are important (\textbf{Fig. 6b-c}).
These include instances of TEAD, FOX and RFX (\textbf{Fig. 6b-c}) correlating with the expression of respectively \textit{TEAD1},
\textit{FOXA2/J3/O1/P1/P2/P4} and \textit{RFX3/4} (\textbf{Fig. 6d} and Supplementary Fig. S7). Instances of a SOX monomer motif is important for one of these topics (\textbf{Fig. 6b-c}), correlating with the expression of \textit{SOX2} (\textbf{Fig. 6d} and Supplementary Fig. S7). Next, in both organoids and embryo cells we identified a topic for which both instances of neuronal progenitor patterns as well as an EBOX pattern is important (organoid Topic 23 and embryo Topic 13 \textbf{Fig. 6b-c}) the importance of this EBOX pattern correlates with the expression of \textit{ASCL1} and \textit{NEUROD4}. For subsequent topics SOX instances are not important anymore (\textbf{Fig. 6b-c}) and we find topics with a where a combination of EBOX and EBF instances are important (organoid and embryo Topic 24; \textbf{Fig. 6b-c}) the latter correlating with the expression of \textit{EBF1-3} (\textbf{Fig. 6d} and Supplementary Fig. S7) and topics where a combination of EBF and ONECUT instances are important (organoid Topic 13 and embryo Topic 18; \textbf{Fig. 6b-c}), the latter correlating with the expression of \textit{ONECUT1-3}. Additionally, in the embryo topics instances of NKX are important in these latter two topics, the importance of these instances is anti-correlated with the expression \textit{NKX2-2} and correlated with the expression of \textit{ISL1} (\textbf{Fig. 6d} and Supplementary Fig. S7). Finally, in the organoids we could identify an additional topic for which instances of GATA are important (\textbf{Fig. 6b-c}) correlating with the expression of \textit{GATA2} (\textbf{Fig. 6d} and Supplementary Fig. S7).\par

@@

@@

In order to create this atlas, we generated a single-cell multiome (combined scATAC-seq and scRNA-seq of the same cell) dataset of the head of a four weeks post conception human embryo and we showed that the embryonic cell states can effectively be reproduced using human neural tube organoids on both transcriptomic level and on the level of the enhancer-codes. Specifically, we identified a total of fifteen distinct cell states with one-to-one matches between the organoid and embryo. The fact that cell states were defined in an unbiased manner by making use of topic modeling and that each of these states were independently identified in two separate datasets increases the confidence of their biological relevance. Furthermore, this high similarity between the organoid and the embyro derived cell states validates neural tube organoids as a biological model for early human neural tube and neural crest development, biological cell states that are otherwise not easily accessible in human.\par

The enhancer code remains complicated to study\cite{wassermanAppliedBioinformaticsIdentification2004} especially
compared to the genetic code\cite{kimDecipheringMultiscaleQuantitative2023, deboerHoldOutGenome2024}. Nonetheless, there
are some recurrent patterns that are shared among enhancers of different cell types. In particular, enhancers underlying
DV patterning, early neuronal fate acquisition and neural crest tend to consist out of clusters of 4-5 different
TFBS\cite{kimDecipheringMultiscaleQuantitative2023, longEverChangingLandscapesTranscriptional2016} that are located
within one NDR (+/-147bp). That is to say, cooperative binding of multiple TFs seem to underlie
the cell type specificity of enhancers\cite{lambertHumanTranscriptionFactors2018}. Here, we see most evidence for a
model of indirect cooperativity\cite{longEverChangingLandscapesTranscriptional2016} (i.e., flexible billboard enhancer
model\cite{arnostiTranscriptionalEnhancersIntelligent2005}). However, for multiple cell types we also observed evidence
of direct coooperativity between TFs in the form of protein-protein interactions between two TFs. This was the case for
TWIST1 and homeodomain TFs heterodimers in facial mesenchyme\cite{kimDNAguidedTranscriptionFactor2024} and GRHL1/2 dimers, TFAP2A and TFAP2B/C  heterodimer\cite{rothsteinHeterodimerizationTFAP2Pioneer2020} and SOX10 homodimers in neural crest. For all these TFs we found dimer motifs with a fixed number nucleotides in between the two TFBS instances. Of note, for SOX10 we observed three different dimer motif with either 3, 4 or 5 bps spacing with those instances with 4 bps spacing being the most accessible (\textbf{Supplementary Fig. 8}). Finally, we also observed evidence of pioneer activity, where binding of one TF can facilitate the subsequent binding of other TFs to the same enhancer\cite{longEverChangingLandscapesTranscriptional2016}. This was the case for FOXA2 in the floorplate, consistent with previous findings of FOXA TFs\cite{iwafuchi-doiPioneerTranscriptionFactor2016, iwafuchiGeneNetworkTransitions2020}. Others have hypothesized, however, that pioneer activity is not a qualitative feature of a TF but rather dependent on the affinity of interactions between TF and DNA\cite{hansenTestPioneerFactor2022}. Similarly, we observed that enhancers that are specific to the floorplate contain either less FOXA2 binding sites and/or binding sites of lower affinity. The use of suboptimal TFBS is a reoccurring feature of cell type specific enhancers\cite{farleySuboptimizationDevelopmentalEnhancers2015, crockerChapterTwentySevenSoft2016, limAffinityoptimizingEnhancerVariants2024, naqviTransferLearningReveals2025}.\par

>>>>

\author[1,2,3,\orcidlink{0000-0001-7907-1247}]{Seppe De Winter}
\author[1,2,3, 4, \orcidlink{0000-0002-1318-2603}]{Camiel Mannens}
\author[1,2,3]{Valerie Christiaens}
\author[4, \orcidlink{0000-0003-1869-0372}]{Lijuan Hu}
\author[5, \orcidlink{0000-0003-2931-8015}]{Erik Sundström}
\author[4, \orcidlink{0000-0002-3491-3444}]{Sten Linnarsson}
\author[1,2,3,*,\orcidlink{0000-0002-8006-0315}]{Stein Aerts}

@@

\affil[1]{VIB Center of AI \& computational Biology, Leuven, Belgium}
\affil[2]{VIB-KU Leuven Center for Brain and Disease Research, Leuven, Belgium}
\affil[3]{KU Leuven Department of Human Genetics, Leuven, Belgium}
\affil[4]{Karolinska Insitute, Department of Biochemistry and Biophysics, Stockholm, Sweden}
\affil[5]{Karolinska Institutet, Division of Neurodegeneration, Department of Neurobiology, Care Sciences and Society,Stockholm, Sweden}

@@

Organoids are a reductionistic biological model system that can be used to interrogate how cell types are encoded in the
genome. Organoids derived from pluripotent stem cells are especially interesting in this context, given that they
demonstrate how a single genome can generate multiple distinct cell types\cite{mccauleyPluripotentStemCellderived2017,
takebeOrganoidsDesign2019, hoferEngineeringOrganoids2021}. Organoids can furthermore give insights into how genomes
interpret both cell intrinsic and extrinsic signals. For example, epigentic
modifications\cite{zenkSinglecellEpigenomicReconstruction2024}, transcription factor (TF)
perturbations\cite{fleckInferringPerturbingCell2023} or morphogen
signaling\cite{sanchis-callejaDecodingMorphogenPatterning2024}. Another benefit of organoids is that they can be used to
model biological systems that are otherwise - due to ethical and/or practical reasons - inaccessible for study. For
example, early human development.
One such system is early human development, the study of which is especially important for our understanding of human health and disease.
\par

@@

Current efforts are ongoing to determine the breadth of human cell types that can be generated using organoid protocols and how they compare to their primary counterparts at a transcriptomic level in for example, brain organoids\cite{heIntegratedTranscriptomicCell2024} and endoderm-derived organoids\cite{xuIntegratedTranscriptomicCell2023}. However, such comparisons at the level of the enhancer-code are still limited\cite{zenkSinglecellEpigenomicReconstruction2024, fleckInferringPerturbingCell2023}. In this study, we will address this gap by determining the enhancer-codes of cell types/states generated by human neural tube organoids, a model system for early human neuronal and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021} and compare these codes to those of cell types/states of the head of a four weeks post conception human embryo.\par

During vertebrate development, both the neural tube and neural crest originate from the neural ectoderm. The neural tube
consists of proliferating neuronal progenitors that, through opposing gradients of SHH and BMP/Wnt signaling, have different identities along the dorso-ventral (DV) axis of the embryo. These early identities later give rise to distinct neuronal subtypes of the central nervous system\cite{sagnerEstablishingNeuronalDiversity2019, delasRepressiveInteractionsGene2020}. Neural crest cells originate from the neural plate border, through intermediate signals of BMP and Wnt\cite{simoes-costaEstablishingNeuralCrest2015}, and migrate throughout the body giving rise to a multitude of cell types including the peripheral nervous system, craniofacial skeleton and melanocytes\cite{simoes-costaEstablishingNeuralCrest2015}. The gene regulatory networks underlying both DV patterning of neuronal progenitors\cite{sagnerEstablishingNeuronalDiversity2019, delasRepressiveInteractionsGene2020} and neural crest induction, specification, migration and differentiation\cite{simoes-costaEstablishingNeuralCrest2015, candido-ferreiraMultilayeredTranscriptionalControl2023} are already well studied. Furthermore, studies using chromatin accessibility profiling methods at the bulk and single cell level, profiling enhancers encoding cell types and states of neuronal progenitors and neural crest cells are starting to emerge\cite{delasDevelopmentalCellFate2023, zhangCisregulatoryLogicIntegrating2024, huEndothelialRegulatoryCircuits2025} but a detailed descriptions of the enhancer-code of individual cell types are still limited.\par

@@

To evaluate the biological relevance of the cell types and states obtained from the organoid, we also performed 10x single

@@

clear that both early differentiating peripheral neurons and to a greater extent early differentiating neurons of the

@@

%TODO:D) The black outline on green text (EMX2/EN1 etc) makes it difficult to read and see the color. consider making it thinner

@@

%TODO: C.M.: I think the original VISTA enhancers were annotated based on localization only, so it would make sense that premig. NC would be very hard to differentiate from normal neural tube ectoderm. Might be good to point out why this overlap exists?

@@

$DeepNeuralTube_{embryo}$ was able to distinguish the different AP domains but grouped topics of similar domains
together (i.e., Topic 59, 31 and 62; Topic 52 and 71), resulting in four distinct AP domains (\textbf{Fig.4 c-d}). By
explaining the predictions of $DeepNeuralTube_{embryo}$ on the regions of those domains we found four distinct
homeodomain TF binding motifs, one for each domain (\textbf{Fig. 4e}). From anterior to posterior this is \textit{TAATTA}, \textit{GGATTA}, \textit{TCATGWTGANTGA} and \textit{CATYMATCA}. Notably, the patterns corresponding to the two posterior most topics correspond to dimer-motifs. Previously it was already hypothesized that genome-wide specific binding of homeodomain TFs is accomplished through cooperative binding of two TFs\cite{mannChapter3Hox2009}. Indeed, the presence of instances of the \textit{TCATGWTGANTGA} motif correlates with the expression of both \textit{PAX5/8} and \textit{EN1/2} and the prescence of instances of the \textit{CATYMATCA} motif correlates with the expression of \textit{MEIS1/2}, \textit{PBX1/2} and \textit{HOXB3} (\textbf{Fig. 4d-e} and Supplementary Fig. S4). In contrast, the patterns corresponding to the two anterior most topics correspond to monomer motifs. The presence of instances of these patterns correlates with the expression of \textit{LHX2/9}, \textit{RAX} and \textit{SIX3/6} for the most anterior topic and with the expression of \textit{OTX1/2}, \textit{EMX2}, \textit{DMBX1} and \textit{LMX1A} for the second most anterior topic (\textbf{Fig. 4d-e} and Supplementary Fig. S4).
%TODO: I'm surprised to see EMX2 (forebrain) together with OTX1/2 and DMBX1 (midbrain)

Focusing on the two most anterior domains, two similar homeodomain TF binding motifs seem to underlie the differential
accessibility of those domains. Both have the core homeobox motif: \textit{TAAT} but one is flanked by the nucleotides
\textit{TAG} while the other is flanked by the nucleotides \textit{CCCT}. 
%TODO: Do we know which TFs these correspond to? If you do, mention it in the text
We were surprised that this slight difference in the two motifs would be enough for encoding two separate AP domains. To test this hypothesis, we first did a substitution experiment where we replaced all instances of \textit{CTAAT\underline{TAG}} with \textit{CTAAT\underline{CCCT}} in regions of the most anterior domain and vice versa (\textbf{Fig. 4f}). Indeed, $DeepNeuralTube_{embryo}$ predicts an anterior-to-posterior shift of accessibility of those regions after the \textit{CTAAT\underline{TAG}} to \textit{CTAAT\underline{CCCT}} substitution and vice versa a posterior-to-anterior shift of accessibility after the \textit{CTAAT\underline{TAG}} with \textit{CTAAT\underline{CCCT}} to  \textit{CTAAT\underline{TAG}} substitution (\textbf{Fig. 4f}). Next, we designed enhancers from scratch by embedding either the  \textit{CTAAT\underline{TAG}} or \textit{CTAAT\underline{CCCT}} motif\cite{taskiranCelltypedirectedDesignSynthetic2024, kempynckCREstedModelingGenomic2025} and using respectively Topic 61 and Topic 31 as target class (\textbf{Fig. 4g}). Embedding of two instances of either motif was sufficient to reach predicted chromatin accessibility levels of equivalent ranges to the predicted chromatin accessibility levels of genomic regions (\textbf{Fig. 4g}). Next, to eliminate potential model bias we repeated the same experiment but optimized for the opposite class. This is, we optimized for Topic 31 by embedding the \textit{CTAAT\underline{TAG}} motif and we optimized for Topic 61 by embedding the \textit{CTAAT\underline{CCCT}} motif (\textbf{Fig. 4h}). In this case, the synthetic design failed to converge for the target class but instead regions with high prediction score for the AP domain that corresponds to the motif used for embedding were generated. This is embedding of the \textit{CTAAT\underline{TAG}} motif still generated regions with high prediction score for Topic 61 and embedding of the \textit{CTAAT\underline{CCCT}} motif generated regions with high prediction score for Topic 31 (\textbf{Fig. 4h}).\par

@@

Focusing on neural crest cells, we identified topics representing cells and regions of both pre- and migratory neural
crest as well as facial mesenchyme (\textbf{Fig. 5a}). Similar to the neuronal progenitors, we used
$DeepNeuralTube_{oranoid}$ and $DeepNeuralTube_{embryo}$ to explain the top 1,000 regions for each organoid and embryo
topic and generated a code-table based on the identified pattern instances (\textbf{Fig. 5b}). Again, both
$DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ found similar patterns and the importance of the instances of
these patterns is highly correlated across both models (0.73 average Pearson correlation coefficient; \textbf{Fig. 5b}).\par

@@

The migratory neural crest cell state is represented by two organoid topics (Topic 65 and 59) and a single embryo topic (Topic 94). Nuclear receptor, AP2, ZIC (only in organoid topics) and SOX-dimer (organoid Topic 59 but not 65) are important for this state (\textbf{Fig. 5b-c}). We could not link the nuclear receptor instances to a specific TF based on the correlation of expression and the importance score of these instances. The importance of AP2 instances correlates with the expression of \textit{TFAP2A}, \textit{TFAP2B} and \textit{TFAP2C} (\textbf{Fig. 5d}), ZIC with \textit{ZIC1} to \textit{ZIC5} and the SOX-dimer with expression of \textit{SOX10} (\textbf{Fig. 5d} and Supplementary Fig. S5). The switch from TFAP2A/C in pre-migratory neural crest to TFAP2A/B in migratory neural crest is consistent with a previously reported function of both heterodimers in respectively neural plate border induction and neural crest specification\cite{rothsteinHeterodimerizationTFAP2Pioneer2020}. After the pre-migratory and before the migratory neural crest state another topic is found in both organoid (Topic 60) and embryo (Topic 105) for which instances of RFX are important (\textbf{Fig. 5b-c}) correlating with the expression of \textit{RFX3/4} (\textbf{Fig. 5d} and Supplementary Fig. S5).\par

@@

Next, we investigated which patterns co-occur in the same genomic regions. For this purpose, we calculated the Jaccard
index of genomic regions based on the presence of an instance of each identified pattern (\textbf{Fig. 5e}). Using this
method we found three sets of regions that are predicted to contain instances of different set of TFs. One set where
instances of TEAD, GRHL and ZEB co-occur, one where instances of Nuclear receptor, AP2 and ZIC
instances co-occur and one set where instances of FOX and TWIST1/homeodomain co-occur.\par
%TODO: C.M.: Mesenchyme+NC or calculated over all regions?

@@

For both the organoid and embryo we found two topics (organoid Topic 6 and 4 and embryo Topic 10 and 18) for which
instances of patterns from neuronal progenitors, more precisely the floorplate, are important (\textbf{Fig. 6b-c}).
These include instances of TEAD, FOX and RFX (\textbf{Fig. 6b-c}) correlating with the expression of respectively \textit{TEAD1},
\textit{FOXA2/J3/O1/P1/P2/P4} and \textit{RFX3/4} (\textbf{Fig. 6d} and Supplementary Fig. S7). Instances of a SOX monomer motif is important for one of these topics (\textbf{Fig. 6b-c}), correlating with the expression of \textit{SOX2} (\textbf{Fig. 6d} and Supplementary Fig. S7). Next, in both organoids and embryo cells we identified a topic for which both instances of neuronal progenitor patterns as well as an EBOX pattern is important (organoid Topic 23 and embryo Topic 13 \textbf{Fig. 6b-c}) the importance of this EBOX pattern correlates with the expression of \textit{ASCL1} and \textit{NEUROD4}. For subsequent topics SOX instances are not important anymore (\textbf{Fig. 6b-c}) instead we find topics where a combination of EBOX and EBF instances are important (organoid and embryo Topic 24; \textbf{Fig. 6b-c}) the latter correlating with the expression of \textit{EBF1-3} (\textbf{Fig. 6d} and Supplementary Fig. S7) and topics where a combination of EBF and ONECUT instances are important (organoid Topic 13 and embryo Topic 18; \textbf{Fig. 6b-c}), the latter correlating with the expression of \textit{ONECUT1-3}. Additionally, in the embryo topics instances of NKX are important in these latter two topics, the importance of these instances is anti-correlated with the expression \textit{NKX2-2} and correlated with the expression of \textit{ISL1} (\textbf{Fig. 6d} and Supplementary Fig. S7). Finally, in the organoids we could identify an additional topic for which instances of GATA are important (\textbf{Fig. 6b-c}) correlating with the expression of \textit{GATA2} (\textbf{Fig. 6d} and Supplementary Fig. S7).\par
%TODO: C.M.: NKX2-2 is an important patterning factor in the neural tube, specifically for ventralization so it should only be expressed around the floor plate. That you don't see it in the organoid might indicate that part of this patterning is not present. https://www.nature.com/articles/19315
%TODO: S.D.W.: I do see it in the DV patterning part, just not in the subset of differentiating neurons.
%TODO: C.M.: That's interesting, so in the organoid it doesn't seem to persist into neuronal differentiation while the embryo retains that NKX2.2 DV patterning? Are all the same types of neurons there or might it be related to a missing cell type?

@@

%TODO: number panels, in A middle, move legend so that text/lines do not class with guiding lines. Color for progenitor and neural crest are very similar and difficult to distinguish. I sometimes use this website to pick colors that are also colorblind friendly. https://colorbrewer2.org/#type=qualitative&scheme=Accent&n=4

@@

In order to create this atlas, we generated a single-cell multiome (combined scATAC-seq and scRNA-seq of the same cell) dataset of the head of a four weeks post conception human embryo and we showed that the embryonic cell states can effectively be reproduced using human neural tube organoids on both transcriptomic level and on the level of the enhancer-codes. Specifically, we identified a total of fifteen distinct cell states with one-to-one matches between the organoid and embryo. The fact that cell states were defined in an unbiased manner by making use of topic modeling and that each of these states were independently identified in two separate datasets increases the confidence of their biological relevance. Furthermore, this high similarity between the organoid and the embyro derived cell states validates neural tube organoids as a biological model for early human neural tube and neural crest development, biological cell states that are otherwise not easily accessible in humans.\par

The enhancer code remains complicated to study\cite{wassermanAppliedBioinformaticsIdentification2004} especially
compared to the genetic code\cite{kimDecipheringMultiscaleQuantitative2023, deboerHoldOutGenome2024}. Nonetheless, there
are some recurrent patterns that are shared among enhancers of different cell types. In particular, enhancers underlying
DV patterning, early neuronal fate acquisition and neural crest tend to consist of clusters of 4-5 different
TFBS\cite{kimDecipheringMultiscaleQuantitative2023, longEverChangingLandscapesTranscriptional2016} that are located
within one NDR (+/-147bp). That is to say, cooperative binding of multiple TFs seem to underlie
the cell type specificity of enhancers\cite{lambertHumanTranscriptionFactors2018}. Here, we see most evidence for a
model of indirect cooperativity\cite{longEverChangingLandscapesTranscriptional2016} (i.e., flexible billboard enhancer
model\cite{arnostiTranscriptionalEnhancersIntelligent2005}). However, for multiple cell types we also observed evidence
of direct coooperativity between TFs in the form of protein-protein interactions between two TFs. This was the case for
TWIST1 and homeodomain TFs heterodimers in facial mesenchyme\cite{kimDNAguidedTranscriptionFactor2024} and GRHL1/2 dimers, TFAP2A and TFAP2B/C  heterodimer\cite{rothsteinHeterodimerizationTFAP2Pioneer2020} and SOX10 homodimers in neural crest. For all these TFs we found dimer motifs with a fixed number nucleotides in between the two TFBS instances. Of note, for SOX10 we observed three different dimer motif with either 3, 4 or 5 bps spacing with those instances with 4 bps spacing being the most accessible (\textbf{Supplementary Fig. 8}). Finally, we also observed evidence of pioneer activity, where binding of one TF can facilitate the subsequent binding of other TFs to the same enhancer\cite{longEverChangingLandscapesTranscriptional2016}. This was the case for FOXA2 in the floorplate, consistent with previous findings of FOXA TFs\cite{iwafuchi-doiPioneerTranscriptionFactor2016, iwafuchiGeneNetworkTransitions2020}. Others have hypothesized, however, that pioneer activity is not a qualitative feature of a TF but rather dependent on the affinity of interactions between TF and DNA\cite{hansenTestPioneerFactor2022}. Similarly, we observed that enhancers that are specific to the floorplate contain either less FOXA2 binding sites and/or binding sites of lower affinity. The use of suboptimal TFBS is a reoccurring feature of cell type specific enhancers\cite{farleySuboptimizationDevelopmentalEnhancers2015, crockerChapterTwentySevenSoft2016, limAffinityoptimizingEnhancerVariants2024, naqviTransferLearningReveals2025}.\par

Our results indicate that enhancers underlying neuronal progenitor AP-patterning form an exception to the rule that enhancers
consist of clusters of 4-5 different TFBS. Homeotypic clusters of 3-4 homeobox instances were sufficient to both create and identify enhancers specific to a certain AP domain. Similar findings were already reported in fly\cite{crockerLowAffinityBinding2015}. It has been a long standing question how homeodomain TFs achieve specific binding\cite{mannChapter3Hox2009, hubertHoxGenesDevelopment2023} given that all homeodomain TFs seem to bind the same \textit{TAAT} motif\cite{bergerVariationHomeodomainDNA2008, noyesAnalysisHomeodomainSpecificities2008}. It has been hypothesized that specificity is mainly achieved through cooperative binding with other TFs\cite{mannChapter3Hox2009}. Indeed, we observe this for EN1/2 and HOXB3 that are predicted to co-bind with PAX5/8 and MEIS1/2 or PBX1/3 respectively forming specific dimer motifs. In contrast, we suggest a novel mode where specificity is achieved through the homeodomain itself for respectively LHX2/9, RAX or SIX3/6 and OTX1/2, EMX2, DMBX1 or LMX1A that preferentially bind \textit{TAATTA} and \textit{TAATCC} sites. Thus, additional nucleotides, next to the core \textit{TAAT}, form a TFBS that is both necessary and sufficient for genome wide cell type specificity. In general, although there are only around 100 different DNA binding domains across human TFs\cite{lambertHumanTranscriptionFactors2018} and thus many of the 1,639 known human TFs recognize similar motifs\cite{lambertHumanTranscriptionFactors2018} there are slight variations in preferential TFBS of TFs of the same family\cite{lambertHumanTranscriptionFactors2018, jolmaDNABindingSpecificitiesHuman2013, dealmeidaDeepSTARRPredictsEnhancer2022} that could be important for cell type specific binding.\par
